---
title: "TBD"
subtitle: "TBD"
author: "Kar Hian Ong and Fadlan Arif"
thanks:  "Code and data are available at: https://github.com/karhian/2020-election"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: |
  | First sentence. Second sentence. Third sentence. Fourth sentence. 
Keywords: "forecasting, US2020 election, Trump, Biden, multilevel regression with post-stratification" 
output:
  bookdown::pdf_document2:
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
```

# Introduction

With the US federal election 2020 swinging by, there is intense debate on who will be the upcoming US president of the United States. The election uses the electoral college system where each state is given a certain number of votes. Candidates with the 270 votes or more will win the election. Us is govern by 2 major parties, Democrats and Republicans. The candidate representing Democrats is Joe Biden while the cendidate representing Republicans is Donald Trump. Both candidates have held positons prior to the 2020 election making them a strong contender for this election. Donald Trump is the current sitting pressident of the United States while Joe Biden was the vice president of the previous administration. 

This reasearch is about who will win the US election. The election result has massive impact for the global population. The sitting president will set the tone for international and domestic policies.    

There are several sections to this paper. Section 2 will talk about the survey data and Acs data, section 3 will detail the model that us use to predict the winner of the US election, section 4 we will get to see the results, section 5 will be discussion on the findings and weaknessess.



# Data

The survey data is provided by Nationscape (@citeNationscape). 

Our data is of penguins (Figure \@ref(fig:bills)).

```{r bills, fig.cap="Bills of penguins", echo = FALSE}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = FALSE) +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

Also bills and their average (Figure \@ref(fig:billssssss)). (Notice how you can change the height and width so they don't take the whole page?)

```{r billssssss, fig.cap="More bills of penguins", echo = FALSE, fig.width=8, fig.height=4}

# This needs to be about the random data tha tI sasmpled?
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = FALSE) +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk way more about it. 



# Model

\begin{equation}
Pr(\theta | y) = \frac{Pr(y | \theta) Pr(\theta)}{Pr(y)}  (\#eq:bayes)
\end{equation}

Equation \@ref(eq:bayes) seems useful, eh?

Here's a dumb example of how to use some references: In paper we run our analysis in `R` [@citeR]. We also use the `tidyverse` which was written by @thereferencecanbewhatever If we were interested in baseball data then @citeLahman could be useful. 


# Results

# Discussion

## First discussion point

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

# Appendix {-}
Code and data are available at: https://github.com/karhian/2020-election

\newpage


# References


